---
title: "Statistical modeling of visual cortical neurons"
output: pdf_document
author: "Cristian Bargiacchi, Christian Mancini"
subject: "Network Analysis"
---

```{r, include=FALSE, eval=FALSE}
library(rmarkdown)
render("NetworkModels.Rmd", output_format = "pdf_document")
```

```{r include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

```{r brasa-tutto, include=FALSE}
rm(list = ls())
```

# Preliminary steps

```{r import-library, message=FALSE}
library(igraph)
library(ergm)
```

We first need to load our data using `igraph`. Data can be found [here](https://github.com/cMancio00/sand-2024/tree/main/Data)

```{r import-data}
neurons_g <- read_graph("Data/mouse_visual.cortex_2.graphml","graphml")
Y = as_adjacency_matrix(neurons_g, sparse = F)
diag(Y) = NA
```

For starting the modeling we first have to convert an igraph object to a network one.

The conversion retains the order of the nodes but we also have to pass the attributes.

```{r data-convertion, message=FALSE, warning=FALSE}
neurons = network(Y, directed = T)
neurons %v% "type1" = vertex_attr(neurons_g,"type1",V(neurons_g))
neurons %v% "type2" = vertex_attr(neurons_g, "type2",V(neurons_g))
```

Now we are good to go!

# Homogeneous Simple Random Graph

Let's start with the simplest model.

**Assumptions**:

-   The probability of forming a tie is the **same** for every pair.

```{r message=FALSE, results=FALSE}
srg_homo = ergm(neurons ~ edges) 
```

```{r}
summary(srg_homo)
```

This model corresponds to a logistic regression, so we can interpret the result as odds:

```{r include=FALSE}
odds = exp(srg_homo$coefficients)
```

> The odds of observing a relation between two randomly selected nodes is about `r sprintf("%.2f", (1-odds)*100)`% lower than that of not observing it.

# Non-Homogeneous Simple Random Graph

**Assumptions**:

-   The same probability of forming a tie is relaxed.
-   Takes in consideration sender and receiver effect

> Since some node do not have in or out degree, those parameter will be set to `-inf`, so the model can't be used, but it can be estimated.

```{r message=FALSE, results=FALSE, eval=FALSE}
srg_no_homo = ergm(neurons ~ edges + sender + receiver,
                   control = control)
```

# Dyad independence model

**Assumptions**:

-   Dyads are independents and follows a *Multinomial* distribution
-   We take in consideration the reciprocity parameter $\gamma_{ij}$ (*mutual*)

## Classic p1 model

**Assumptions**:

-   the reciprocity parameter is $\gamma =\gamma_{ij}$
-   $\mu_{ij}$ depends additively from on the sender and receiver effect of node $i$ and $j$ involved.

```{r message=FALSE, warning=FALSE, results=FALSE, eval=FALSE}
p1_classic = ergm(neurons ~ edges + sender + receiver + mutual,
                  control = control.ergm(seed = 1, main.method = "Stochastic-Approximation"))
```

## Sender and reciver independency assumption

We now construct 3 p1 model to check for *reciprocity* with the following assumptions:

1.  Sender effect independent
2.  Receiver effect independent
3.  Sender and receiver effect independent (equals to non-homogeneous SRG)

> For the same reasons explained in the non-homogeneous SRG, these models can't be estimated. The coefficients are set to `-inf`.

```{r message=FALSE, results=FALSE, eval=FALSE}
p1_sender_ind = ergm(neurons ~ edges + receiver + mutual,
                     control = control.ergm(seed = 1))
```

```{r message=FALSE, results=FALSE, eval=FALSE}
p1_receiver_ind = ergm(neurons ~ edges + sender + mutual,
                       control = control.ergm(seed = 1))
```

```{r message=FALSE, results=FALSE, eval=FALSE}
p1_mutual_only = ergm(neurons ~ edges + mutual,
                      control = control.ergm(seed = 1))
```

# Nodal attributes

In this part of the analysis we include nodal attributes to explore *homophily* and *main* effects.

-   **Main effect**: Nodes of a specific type have more chance to form ties

-   **Homophily effect** Nodes of a specific type have more chance to form ties between nodes of the same type

> As saw in the descriptive analysis we expect to observe dissortative mixing.

Since we only have categorical attributes we will use:

-   `nodefactor()` to include main effect
-   `nodematch()` to include homophily effect

Let's explore "*type1*" nodal attribute.

```{r message=FALSE, results=FALSE}
main_homo_type_one = ergm(neurons ~ edges + nodefactor("type1") + nodematch("type1"), 
            control = control.ergm(seed = 1)) 

```

```{r}
summary(main_homo_type_one)
```

As we expect, the *Dissortative mixing* is captured. The probability of forming ties with nodes of the same type is `r sprintf("%.2f", (1-(exp(-4.10482)))*100)`% lower, with respect to probability of forming ties with different one. Moreover, "*Characterized pyramidal neuron*" have more chances to form ties (which is true because they start the synapses). More precisely, the odds is `r sprintf("%.2f", exp(4.12940))` times higher respect to a "*Cell body in EM volume*".

Now we use *"type2"* attribute, but only as main effect, and we remove `mutual` since it is estimated as `-inf`.

> We exclude homophily, since all the coefficents will result non significants.

```{r message=FALSE, results=FALSE}
main_type_two = ergm(neurons ~ edges + nodefactor("type2"),
            control = control.ergm(seed = 1)) 
```

```{r}
summary(main_type_two)
```

We basically got the same conclusions. Since the reference is "*NA*" which can only be "*Characterized pyramidal neuron*", the odds of form a tie are smaller if a node is "*excitatory*" or "*inhibitory*".

Let's now put all together:

```{r message=FALSE, results=FALSE}
main_homo = ergm(neurons ~ edges + nodefactor("type1") + nodefactor("type2") 
                 + nodematch("type1"), 
            control = control.ergm(seed = 1)) 
```
```{r}
summary(main_homo)
```
We can see that the main effect of being a *Dendritic fragment* is not significant, so the tendency to form more ties than by chance depends only of being a *Characterized pyramidal neuron* or not.

We can now compare BIC for these models including nodal attributes.

```{r include=FALSE}
bic_main_homo_type_one <- sprintf("%.2f", BIC(main_homo_type_one))
bic_main_type_two <- sprintf("%.2f", BIC(main_type_two))
bic_main_homo <- sprintf("%.2f", BIC(main_homo))
```

| Model              | BIC                        |
|--------------------|----------------------------|
| Main Homo Type One | `r bic_main_homo_type_one` |
| Main Type Two      | `r bic_main_type_two`      |
| Main Homo          | `r bic_main_homo`          |

According to `BIC` the full model is better.

# Nodal attributes (Binary)

Let's repeat the above analysis but using new attributes which are the binarization of the real ones.

As reference category we choose "*Characterized pyramidal neuron*" and "*Postsynaptic excitatory target*".

Generate the new attributes:

```{r message=FALSE, results=FALSE}
type1.new = rep(0, 195)
type1.new[vertex_attr(neurons_g,"type1") == "Characterized pyramidal neuron"] = 1
type1.new
neurons %v% "type1.new" = type1.new

type2.new = rep(0, 195)
type2.new[vertex_attr(neurons_g,"type2") == "Postsynaptic excitatory target"] = 1
type2.new
neurons %v% "type2.new" = type2.new
```

Estimate all the models again:

```{r message=FALSE, results=FALSE}
main_homo_type_one_binary = ergm(neurons ~ edges + nodefactor("type1.new") 
                                 + nodematch("type1.new"), 
            control = control.ergm(seed = 1))

main_type_two_binary = ergm(neurons ~ edges + nodefactor("type2.new"),
            control = control.ergm(seed = 1))

main_homo_binary = ergm(neurons ~ edges + nodefactor("type1.new") 
                        + nodefactor("type2.new") + nodematch("type1.new"), 
            control = control.ergm(seed = 1)) 
```

```{r include=FALSE}
bic_main_homo_type_one_binary <- sprintf("%.2f", BIC(main_homo_type_one_binary))
bic_main_type_two_binary <- sprintf("%.2f", BIC(main_type_two_binary))
bic_main_homo_binary <- sprintf("%.2f", BIC(main_homo_binary))
```

| Model                     | BIC                               |
|---------------------------|-----------------------------------|
| main_homo_type_one_binary | `r bic_main_homo_type_one_binary` |
| main_type_two_binary      | `r bic_main_type_two_binary`      |
| main_homo_binary          | `r bic_main_homo_binary`          |

According to `BIC` the new model with just "*type1.new*" is the best for now with a score of `r sprintf("%.2f", BIC(main_homo_type_one_binary))`.

Let's explore the summary of the model

```{r}
summary(main_homo_type_one_binary)
```

As we can see every parameter is significant. The interpretation is the same as before.

# Markov

si piglia il modello bono e si aggiunge roba, ostar(2) non va --> stelle K-alternati dopo 



```{r message=FALSE, warning=FALSE}
markov = ergm(neurons ~ edges + nodefactor("type1.new") + nodematch("type1.new") + istar(2) + triangles, 
             control = control.ergm(seed = 1))  
```
```{r}
summary(mod9)
```

```{r message=FALSE, warning=FALSE}
markov_no_triangles = ergm(neurons ~ edges + nodefactor("type1.new") 
                           + nodematch("type1.new") + nodefactor("type2.new") + istar(2), 
             control = control.ergm(seed = 1))
```

```{r}
summary(markov_no_triangles)
```

 metti bic di markov.
 
 
# Markov k-star alternate

Questi mi servono perché ostar non va, quindi uso i k-star alternati


```{r message=FALSE, warning=FALSE}
k_star = ergm(neurons ~ edges + nodefactor("type1.new") + nodefactor("type2.new") + nodematch("type1.new") + gwidegree(decay = 1, fixed = TRUE) + gwodegree(decay = 1, fixed = TRUE) ,
             control = control.ergm(seed = 1, main.method = "Stochastic-Approximation"))
```
```{r}
summary(k_star)
```

# Social Circuits

```{r message=FALSE, warning=FALSE}
social = ergm(neurons ~ edges + nodefactor("type1.new") + nodefactor("type2.new") 
             + nodematch("type1.new") + 
                gwdsp(decay = 1, fixed = T), control = control.ergm(seed=1, main.method = "Stochastic-Approximation" ))
```
```{r}
summary(social)
```

I meglio sono Markov_no_triangoli e k-star

# Simulazione

```{r}
sim_k_star = simulate(k_star, nsim = 100, verbose = TRUE, seed = 1)
```
```{r}
library(intergraph)

fnc = function(xx){
  ig = asIgraph(xx)
  tr = transitivity(ig)
  ideg = sd(degree(ig, mode = "in"))
  odeg = sd(degree(ig, mode = "out"))
  return(c(tr, ideg, odeg))
}

null.distr = matrix(,100,3)
for(b in 1:100){
  null.distr[b,]  = fnc(sim_markov_no_triangles[[b]])
}
dev.new()
par(mfrow = c(3,1))
hist(unlist(null.distr[,1]), xlab = "transitivity"); abline(v = transitivity(neurons_g), col = "red")
hist(unlist(null.distr[,2]), xlab = "in-degree"); abline(v = sd(degree(neurons_g, mode = "in")), col = "red")
hist(unlist(null.distr[,3]), xlim = c(3.5,5.25),xlab = "out-degree"); abline(v = sd(degree(neurons_g, mode = "out")), col = "red")

```
